### 2024-03-29 parts/intro.tex
correct citations
rewrote intro

\AP There have been recurrent attempts to generalize the notions of regularity
from languages to functions, leading to classes of various expressiveness (for
instance, sequential, rational, regular, and \kl{polyregular functions})
\cite{ELME65,AHUL69,EILE74,ENMA02}. In this paper, we will focus on
\intro{polyregular functions} ($\intro*\Poly$), that shares with regular
languages its characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers
\cite{BOKL19}. As for regular languages, the class $\intro*\SF$ of
\intro{star-free polyregular function} can equivalently be introduced in terms
of first-order definability, counter-free pebble transducers, and suitable
restrictions of for-tranducers.

\AP
Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton. For
sequential functions (that admit a canonical object), aperiodicity is decidable
\cite{CHOF03}, and more recently, aperiodicity has been proven decidable
for rational functions \cite{FGL16,FGLM18}. The
problem remains open for regular functions, although it has been proven
decidable when changing the semantics of function equality to account for the
\emph{origins} in the output \cite{BOJA14}. To construct subclasses of
\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
of a function $f$, which is given by a number $k \in \Nat$, which can be used
to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
controls the maximal complexity of the $\MSO$ interpretation needed to
represent the function \cite{BOJA22}.

\AP
Another way to approach the problem of deciding if a \kl{polyregular function}
is \kl{star-free}, is to not restrict the computational power (via the
\kl{growth rate}), nor change the semantics (by adding origins), but consider
restrictions of the output alphabet. In particular, there have been successful
results proven on \emph{unary polyregular functions}, also known as
\kl{$\Nat$-polyregular functions} ($\intro*\NPoly$)
\cite{DOUE21,DOUE22}. For instance, the \kl{growth
rate} $k$ of a function $f \in \NPoly$ provides the minimal number of pebbles
needed to compute $f$ \cite{DOUE21}, which does not hold on $\Poly$
\cite{BOJA22}. Let us remark that the restriction to
\emph{unary output} in the case of \kl{polyregular functions} can be lifted to
outputs in the free commuative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{unary output} is really about
enforcing commutativity on the output.

\AP
Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
\cite{CDTL23}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). Although the proof method does not carry
from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
that the semantic characterization of \kl{star-free $\Rel$-polyregular
functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$).

\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
    \label{zsf-nsf:conjecture}
    Let $k \in \Nat$.
    A function $f \in \NPoly[k]$
    belongs to $\NSF[k]$ if and only if
    it is \kl{ultimately polynomial} (defined page \kpageref{ultimately polynomial}).
    In particular,
    $\NSF[k] = \ZSF[k] \cap \NPoly$.
\end{conjecture}

\AP
It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
ambiguity, and a notion of \emph{quantitative} $\MSO$
\cite{SCHU62,KRRC13}, and therefore can benefit from generic
theorems obtained on these \cite{BERE10}. However, for these
larger classes of functions, several competing notions of aperiodicity exist
\cite{DRGA19,REUT80,CDTL23}, and the
decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved by
\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
\cite[Open question 5.55]{DOUE23}.

\AP
We refer the reader to \cref{previously-known-inclusions:fig} to get an
overview of the different classes of functions appearing in the paper. Arrows
denote strict inclusions, and effectiveness (both in terms of decidability and
of effective representation) is represented by thick double arrows. Inclusions
that are suspected to be effective are represented using a dashed arrow
together with a question mark.

@@

\paragraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
functions} that are furthermore \kl{commutative}. Remark that this enforces a
commutativity property both on the input alphabet and on the output alphabet.
In this restricted setting, we answer positively to \cref{zsf-nsf:conjecture},
and prove that the membership problems between $\NSF$, $\ZSF$, $\NPoly$, and
$\ZPoly$ are all decidable, with effective conversion algorithms. This
effectively shows that all arrows in the lower left square of
\cref{previously-known-inclusions:fig} are decidable in the \kl{commutative}
setting. Furthermore, we characterize \kl{commutative} \kl{$\Nat$-polyregular
function} as the class of \kl{$\Rel$-polyregular functions} that are
\kl{combinatorial} (defined page \kpageref{combinatorial}), which is a semantic
property akin to the one \kl{ultimate polynomiality} that characterizes $\ZSF$.
As a byproduct of our analysis, we provide a counter example to the
characterization of polynomials computed by \kl{$\Nat$-rational series} given
by \textcite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.

@@

\paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
the multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
$\Rel$-polyregular functions), show that one can decide if a function $f \in
\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give
the basic definitions used on multivariate polynomials. In
\cref{polynomials:sec}, we focus on \emph{polynomials}, by introducing the
\cref{karh:thm} of \textcite{KARH77}, providing a counter-example in
\cref{thm:counter-example}, and an alternative characterization in
\cref{corrected-version:thm}. In \cref{beyond-polynomials:sec}, we leverage the
results over polynomials to decide $\NPoly$ inside $\ZPoly$ in the
\kl{commutative} case (\cref{decidable-n-poly:thm}), and prove that
\cref{zsf-nsf:conjecture} holds for \kl{commutative} functions
(\cref{zsf-npoly-nsf:thm}). Finally, in \cref{beyond-commutative:sec}, we
introduce the notion of \kl{residual transducer}
(\cref{residual-transducer:def}), and prove in \cref{non-commutative-npoly:thm}
that it characterizes $\NPoly$ inside $\ZPoly$ through a notion of
\kl{well-quasi-ordered} collection of residuals.

>>>>

\AP There have been recurrent attempts to generalize the notions of regularity
from languages to functions, leading to classes of various expressiveness:
Mealy Machines ($\intro*\Mealy$) \cite{MEAL55}, sequential functions
($\intro*\Sequential$) \cite{SCHU77}, rational functions ($\intro*\Rational$)
\cite{EILE74}, regular functions ($\intro*\Regular$) \cite{ENMA02},
and polyregular functions ($\intro*\Poly$) \cite{BOKL19}. For all of these
models, an \emph{aperiodic} counterpart can be defined (respectively,
$\intro*\AMealy$, $\intro*\ASequential$, $\intro*\ARational$,
$\intro*\ARegular$, $\intro*\SF$), lifting the correspondence between
counter-free automata, star-free languages and aperiodic monoids to the
functional setting \cite{FKT14,BOJA14,CADA15,BOKL19}. The inclusions between
these classes of functions are all known to be strict, and we depicted in
\cref{computational-models:fig} the status of the \emph{membership problem}
associated to these strict inclusions (that is, given a function $f$, decide if
$f$ can be computed by a function of a proper subclass). Note that in
\cref{computational-models:fig}, if $\mathsf{B}$ is a proper subclass of
$\mathsf{C}$, then the \emph{aperiodic} variant of $\mathsf{B}$ is precisely
the intersection of $\mathsf{B}$ with the \emph{aperiodic} variant of the
larger class $\mathsf{C}$. As an example, to conclude that a function can be
computed by an \emph{aperiodic sequential function} ($\ASequential$) it is
enough to prove that it is sequential ($\Sequential$), and computed by an
aperiodic polyregular function ($\SF$).

\AP To approach the \emph{aperiodicity} membership problems the traditional
approach has been to develop so-called canonical objects, akin to the minimal
automaton or the syntactic monoid of regular languages, over which aperiodicity
is equivalent to a simple syntactic criterion (e.g. having no counters). This
is the proof scheme that has been implemented for sequential functions
\cite{CHOF03}, and more recently for rational functions \cite{FGL18}.
Unfortunately, such a canonical object remains to be defined for regular and
polyregular functions.

\begin{figure}
    \centering
    \begin{tikzpicture}[
        xscale=1.2,
        decidable/.style={
            ->, thick, color=black
        },
        unknown/.style={
            ->, dashed, color=gray
        }
        ]
        % Create nodes for the classes and their aperiodic variants
        % Mealy / sequential / rational / regular / polyregular
        \node (mealy)  at (0, 2) {$\Mealy$};
        \node (seq)    at (2, 2) {$\Sequential$};
        \node (rat)    at (4, 2) {$\Rational$};
        \node (reg)    at (6, 2) {$\Regular$};
        \node (poly)   at (8, 2) {$\Poly$};
        % Aperiodic variants
        \node (amealy) at (0,0) {$\AMealy$};
        \node (aseq)   at (2,0) {$\ASequential$};
        \node (arat)   at (4,0) {$\ARational$};
        \node (areg)   at (6,0) {$\ARegular$};
        \node (apoly)  at (8,0) {$\SF$};
        % Arrows
        % The decidability is known for inclusions
        % mealy -> sequential -> rational -> regular -> polyregular
        % similarly for the aperiodic variants
        \draw[decidable] (mealy) --
            node[midway, above] {folklore}
            (seq);
        \draw[decidable] (seq)   -- 
            node[midway, above] {\cite{CHOF03}}
            (rat);
        \draw[decidable] (rat)   -- 
            node[midway, above] {\cite{FGRS13,BGMP15}}
            (reg);
        \draw[decidable] (reg)   -- 
            node[midway, above] {\cite{BOKL19}}
            (poly);
        \draw[decidable] (amealy) --
            (aseq);
        \draw[decidable] (aseq) --
            (arat);
        \draw[decidable] (arat) --
            (areg);
        \draw[decidable] (areg) --
            (apoly);
        % Decidability of the aperidoic variants are known
        % for mealy, sequential, rational but not for regular nor polyregular.
        % we draw the arrows
        \draw[decidable] (amealy) --
            node[midway, left] {\cite{SCHU65,MNPA71}}
            (mealy);
        \draw[decidable] (aseq) -- 
            node[midway, left] {\cite{CHOF03}}
            (seq);
        \draw[decidable] (arat) -- 
            node[midway, left] {\cite{FGL16,FGLM18}}
            (rat);
        \draw[unknown] (areg) --
            node[midway, left] {$\simeq$\cite{BOJA14}}
            (reg);
        \draw[unknown] (apoly) --
            node[midway, left] {?}
            (poly);
    \end{tikzpicture}
    \caption{
        In this picture, arrows denote strict inclusions. Plain
        arrows signify that the corresponding membership problem is decidable,
        while dashed arrows signify that the corresponding membership problem
        is conjectured to be decidable. In the case of \kl{regular functions}
        ($\Regular$),
        the decidability has only been obtained under a different notion
        of equivalence of functions \cite{BOJA14}.
    }
    \label{computational-models:fig}
\end{figure}

\AP In order to approach the decidability of decision problems on
\kl{polyregular functions}, restricting the output alphabet to a single letter
has proven to be a fruitful approach \cite{DOUE21,DOUE22}. The key advantage of
single-letter output alphabets is that the particular order in which letters
are produced by the machine is irrelevant. This allows to obtain results that
would otherwise fail because of this ordering, such as the so-called
\emph{pebble minimization theorem} of \cite{DOUE21}, which fails in the general
case \cite{BOJA22,KLEP23}. Let us remark that the restriction to \emph{unary
output} in the case of \kl{polyregular functions} can often be lifted to
outputs in the free commutative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{single letter output alphabet} is
really about enforcing commutativity on the output. However, even with the
restriction to single letter output alphabets, the decidability of
\kl{star-free polyregular functions} among \kl{polyregular functions} remains
an open problem \cite[Conjecture 7.61]{DOUE23}.

\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
as having outputs in $\Nat$, and this is why they were introduced respectively
as \kl{$\Nat$-polyregular functions} ($\intro*\NPoly$) and \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$) \cite{DOUE21,DOUE22}. From this
point of view, $\NPoly$ is a subclass of a larger and well-studied class of
functions called (non-commutative) \intro{rational series}
\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
\kl{$\Nat$-polyregular functions} correspond to a natural subclass of
\kl{$\Nat$-rational series} that was frequently rediscovered in the literature
\cite{SCHU62,KRRC13}. In \cite{CDTL23}, the authors introduced the class of
\kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$) that generalize
\kl{$\Nat$-polyregular functions} by allowing negative outputs. This remains a
subclass of \kl{rational series}, more precisely, of the class of
\kl{$\Rel$-rational series} ($\intro*\ZRat$). For this class of functions the
aperiodic variant $\intro*\ZSF$ has been shown to be decidable \cite{CDTL23}.
Note however that $\ZPoly$ does not correspond to a restriction of
\kl{polyregular functions}, because of the extra expressive power introduced by
the negative outputs \cite{CDTL23}. Furthermore, deciding the membership of
$\NRat$ inside $\ZRat$ is an open problem
(see e.g. \cite{KARH77}),
that remains open even in the
restricted case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}.

\AP It was conjectured that, analogously to classes appearing in
\cref{computational-models:fig}, the following equation would hold $\NSF =
\NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}. This would imply the
decidability of aperiodicity for single letter output polyregular functions
since aperiodicity is decidable for $\ZSF$. We refer the reader to
\cref{previously-known-inclusions:fig} to get an overview of the different
classes of functions appearing in the paper. Arrows denote strict inclusions,
and effectiveness (both in terms of decidability and of effective
representation) is represented by thick double arrows. Inclusions that are
suspected to be effective are represented using a dashed arrow together with a
question mark.

@@

\paragraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
functions} that are furthermore \kl{commutative}. Remark that this enforces a
commutativity property both on the input alphabet and on the output alphabet.
In this restricted setting, we answer positively to \cref{zsf-nsf:conjecture},
and prove that the membership problems between $\NSF$, $\ZSF$, $\NPoly$, and
$\ZPoly$ are all decidable, with effective conversion algorithms. This
effectively shows that all arrows in the lower left square of
\cref{previously-known-inclusions:fig} are decidable in the \kl{commutative}
setting. Furthermore, we characterize \kl{commutative} \kl{$\Nat$-polyregular
function} as the class of \kl{$\Rel$-polyregular functions} that are
\kl{combinatorial} (defined page \kpageref{combinatorial}), which is a semantic
property akin to the one \kl{ultimate polynomiality} that characterizes $\ZSF$.
As a byproduct of our analysis, we provide a counter example to the
characterization of polynomials computed by \kl{$\Nat$-rational series} given
by \cite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.

@@

\paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
the multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
$\Rel$-polyregular functions), show that one can decide if a function $f \in
\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give
the basic definitions used on multivariate polynomials. In
\cref{polynomials:sec}, we focus on \emph{polynomials}, by introducing the
\cref{karh:thm} of \cite{KARH77}, providing a counter-example in
\cref{thm:counter-example}, and an alternative characterization in
\cref{corrected-version:thm}. In \cref{beyond-polynomials:sec}, we leverage the
results over polynomials to decide $\NPoly$ inside $\ZPoly$ in the
\kl{commutative} case (\cref{decidable-n-poly:thm}), and prove that
\cref{zsf-nsf:conjecture} holds for \kl{commutative} functions
(\cref{zsf-npoly-nsf:thm}). Finally, in \cref{beyond-commutative:sec}, we
introduce the notion of \kl{residual transducer}
(\cref{residual-transducer:def}), and prove in \cref{non-commutative-npoly:thm}
that it characterizes $\NPoly$ inside $\ZPoly$ through a notion of
\kl{well-quasi-ordered} collection of residuals.

### 2024-03-29 parts/prelim.tex
rewrote intro

\AP In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of
integers (resp. nonnegative integers). The capital letters $\Sigma,\Gamma$
denotes fixed alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$
(resp. $\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
$\Sigma, \Gamma$. The empty word is written $\varepsilon \in \Sigma^*$. When $w
\in \Sigma^*$ and $a \in \Sigma$, we let $\card{w} \in \Nat$ be the length of
$w$, and $\card[a]{w}$ be the number of occurrences of $a$ in $w$. We assume
that the reader is familiar with the basics of automata theory, in particular
the notions of monoid morphisms, idempotents in monoids, monadic second-order
($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
\cite{thomas1997languages}).

@@

\AP Because the functions of interest in this paper have output in $\Nat$ or
$\Rel$, we will only provide the definiton of \kl{polyregular functions} for
these two output semgroups, and we refer the reader to
\cite{bojanczyk2019string} for the general definition of \kl{polyregular
functions}. The following \cref{nat-rel-poly:def} is one of the equivalent
definitions of \cite{LOPEZ23b}, and is similar in shape to the \emph{finite
counting automata} introduced by \textcite{schutzenberger1962}. Let us recall
that a monoid $M$ is \intro(monoid){aperiodic} whenever there exists $n \in
\Nat$ such that for all $x \in M$, $x^{n+1} = x^n$.

\begin{definition}[$\Rel$-polyregular functions {\cite{LOPEZ23b}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a production function.
    The \intro{$\Rel$-polyregular function}
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is computed as follows:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is called \intro{$\Nat$-polyregular}.
    When the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \intro{star-free $\Rel$-polyregular}
    (resp. \intro{star-free $\Nat$-polyregular}).
\end{definition}

\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF$.
\end{example}

\AP Let us briefly recall in the following lemma alternative characterizations
of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
analysis of their \kl{commutative} counterpart, and others that provide a nicer
syntax to construct examples. These characterizations were previously known,
and can be found for instance in \cite{gaetanphd,LOPEZ23b}.

\begin{lemma}
    \label{polynomial-rational-polyreg:fact}
    Let $f \in \NRat$ (resp. $f \in \ZRat$), then
    the following are equivalent:
    \begin{enumerate}
        \item $f \in \NPoly$ (resp. $\ZPoly$);
        \item $f$ is a \kl{polyregular function} with output
            in $\set{1}^*$,
            postcomposed with $\polysum$
            (resp. with output in $\set{-1,+1}^*$);
        \item $f$ has \kl{polynomial growth};
        \item \label{npoly-counting-mso:item} $f$ belongs to
            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \MSO})$
            with $\mathbb{S} = \Nat$
            (resp. $\mathbb{S} = \Rel$).
    \end{enumerate}
\end{lemma}

@@

\begin{lemma}
    Let $f \in \NPoly$ (resp. $f \in \ZPoly$), then
    the following are equivalent:
    \begin{enumerate}
        \item $f \in \NSF$ (resp. $\ZSF$);
        \item $f$ is a \kl{star-free polyregular function} with output
            in $\set{1}^*$,
            postcomposed with $\polysum$
            (resp. with output in $\set{-1,+1}^*$);
        \item $f$ belongs to
            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \FO})$
            with $\mathbb{S} = \Nat$
            (resp. $\mathbb{S} = \Rel$).
    \end{enumerate}
\end{lemma}

@@

\AP Inside $\ZPoly$ there is an infinite hierarchy of classes, defined by
limiting the maximal number of free variables used to construct the function as
a sum of counting formulas (\cref{npoly-counting-mso:item}). We write
$\ZPoly[k]$ (resp. $\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) for the functions in
$\ZPoly$ (resp. $\NPoly$, $\ZSF$, $\NSF$) that can be constructed using at most
$k$ free variables. It has been proven in
\cite{LOPEZ23b,bojanczyk2022transducers} that the classes $\ZPoly[k]$ (resp.
$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are characterized by the \kl{growth rate},
that is, $f \in \ZPoly[k]$ if and only if $f \in \ZPoly$ and $\card{f(w)} =
\bigO\left(\card{w}^k\right)$. and similarly for the other classes.

@@

\subsection{Polynomials} \AP All polynomials considered in this paper have
coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
\Rel[X_1, \dots, X_n]$ is a \intro{$\Nat$-rational polynomial} if it is
\kl{represented} by a \kl{$\Nat$-polyregular function}. The analogue notion of
\intro{$\Rel$-rational polynomials} is not of particular interest, since every
polynomial $P$ is \kl{represented} by a \kl{$\Rel$-polyregular function}.

\begin{example}
    \label{negative-not-nrat:ex}
    The polynomials $X$, and $X^2 + 3$ are \kl{$\Nat$-rational polynomials},
    but $- X$ is a \kl{$\Rel$-rational polynomial} that is 
    not a \kl{$\Nat$-rational polynomial}.
\end{example}

@@

\begin{fact}
    \label{pre-image-regular:fact}
    The pre-image of a regular language by a \kl{$\Nat$-rational series}
    is a regular language. 
\end{fact}

\begin{example}
    Let $P(X, Y) \defined (X - Y)^2$.
    Then $P$ is
    is \kl{non-negative}, but is
    not a \kl{$\Nat$-rational polynomial}.
\end{example}
\begin{proof}
    Assume by contradiction that
    $f \in \NPoly$ \kl{represents} $f$ over the alphabet $\Sigma \defined \set{a,b}$.
    Then, $f^{-1}(\set{0})$ is a regular language
    (\cref{pre-image-regular:fact}),
    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
    is not.
\end{proof}

>>>>

\AP In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of
integers (resp. non-negative integers). The capital letters $\Sigma,\Gamma$
denotes fixed alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$
(resp. $\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
$\Sigma, \Gamma$. The empty word is written $\varepsilon \in \Sigma^*$. When $w
\in \Sigma^*$ and $a \in \Sigma$, we let $\card{w} \in \Nat$ be the length of
$w$, and $\card[a]{w}$ be the number of occurrences of $a$ in $w$. We assume
that the reader is familiar with the basics of automata theory, in particular
the notions of monoid morphisms, idempotents in monoids, monadic second-order
($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
\cite{THOM97}).

@@

\AP
% How the problem has been approached so far
To construct subclasses of
\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
of a function $f$, which is given by a number $k \in \Nat$, which can be used
to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
controls the maximal complexity of the $\MSO$ interpretation needed to
represent the function \cite{BOJA22}.

\AP Because the functions of interest in this paper have output in $\Nat$ or
$\Rel$, we will only provide the definition of \kl{polyregular functions} for
these two output semigroups, and we refer the reader to \cite{BOKL19} for the
general definition of \kl{polyregular functions}. The following
\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
and is similar in shape to the \emph{finite counting automata} introduced by
\cite{SCHU62}. Let us recall that a monoid $M$ is \intro(monoid){aperiodic}
whenever for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} =
x^n$.

\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a production function.
    The \intro{$\Rel$-polyregular function}
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is computed as follows:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is called \intro{$\Nat$-polyregular}.
    When the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \intro{star-free $\Rel$-polyregular}
    (resp. \intro{star-free $\Nat$-polyregular}).
\end{definition}

\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF$.
    \proofref{size-of-word-nsf:ex}
\end{example}

\AP Let us briefly recall in the following lemma alternative characterizations
of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
analysis of their \kl{commutative} counterpart, and others that provide a nicer
syntax to construct examples. These characterizations were previously known,
and can be found for instance in \cite{DOUE23,CDTL23}.

\begin{lemma}
    \label{polynomial-rational-polyreg:fact}
    Let $f \in \NRat$ (resp. $f \in \ZRat$), then
    the following are equivalent:
    \begin{enumerate}
        \item $f \in \NPoly$ (resp. $\ZPoly$);
        \item $f$ is a \kl{polyregular function} with output
            in $\set{1}^*$,
            post-composed with $\polysum$
            (resp. with output in $\set{-1,+1}^*$);
        \item $f$ has \kl{polynomial growth};
        \item \label{npoly-counting-mso:item} $f$ belongs to
            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \MSO})$
            with $\mathbb{S} = \Nat$
            (resp. $\mathbb{S} = \Rel$).
    \end{enumerate}
\end{lemma}

@@

\begin{lemma}
    Let $f \in \NPoly$ (resp. $f \in \ZPoly$), then
    the following are equivalent:
    \begin{enumerate}
        \item $f \in \NSF$ (resp. $\ZSF$);
        \item $f$ is a \kl{star-free polyregular function} with output
            in $\set{1}^*$,
            post-composed with $\polysum$
            (resp. with output in $\set{-1,+1}^*$);
        \item $f$ belongs to
            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \FO})$
            with $\mathbb{S} = \Nat$
            (resp. $\mathbb{S} = \Rel$).
    \end{enumerate}
\end{lemma}

@@

\AP Inside $\ZPoly$ there is an infinite hierarchy of classes, defined by
limiting the maximal number of free variables used to construct the function as
a sum of counting formulas (\cref{npoly-counting-mso:item}). We write
$\ZPoly[k]$ (resp. $\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) for the functions in
$\ZPoly$ (resp. $\NPoly$, $\ZSF$, $\NSF$) that can be constructed using at most
$k$ free variables. It has been proven in
\cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are characterized by the \kl{growth rate},
that is, $f \in \ZPoly[k]$ if and only if $f \in \ZPoly$ and $\card{f(w)} =
\bigO\left(\card{w}^k\right)$, and similarly for the other classes.

@@

\subsection{Polynomials} \AP All polynomials considered in this paper have
coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
\Rel[X_1, \dots, X_n]$ is an \intro{$\Nat$-rational polynomial} if it is
\kl{represented} by a \kl{$\Nat$-polyregular function}. The analogue notion of
\intro{$\Rel$-rational polynomials} is not of particular interest, since every
polynomial $P$ is \kl{represented} by a \kl{$\Rel$-polyregular function}.

\begin{example}
    \label{negative-not-nrat:ex}
    The polynomials $X$, and $X^2 + 3$ are \kl{$\Nat$-rational polynomials},
    but $- X$ is a \kl{$\Rel$-rational polynomial} that is 
    not an \kl{$\Nat$-rational polynomial}.
\end{example}

@@

\begin{fact}
    \label{pre-image-regular:fact}
    The pre-image of a regular language by an \kl{$\Nat$-rational series}
    is a regular language. 
\end{fact}

\begin{example}
    Let $P(X, Y) \defined (X - Y)^2$.
    Then $P$ is
    is \kl{non-negative}, but is
    not an \kl{$\Nat$-rational polynomial}.
\end{example}
\begin{proof}
    Assume by contradiction that
    $f \in \NPoly$ \kl{represents} $f$ over the alphabet $\Sigma \defined \set{a,b}$.
    Then, $f^{-1}(\set{0})$ is a regular language
    (\cref{pre-image-regular:fact}),
    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
    is not.
\end{proof}
